---
title: "Untitled"
output: html_document
date: "2023-05-25"
---

```{r setup, results='hide', message=FALSE}
library(readxl)
library(tidyverse)
library(gdata)
library(car)
```

The first step is to load in the data from the Excel file and combine it with the metadata.

```{r}
data_raw <- read_excel("data/arabidopsis_data.xlsx", col_types = c("text", rep("numeric", 8), "text")) #making sure every value is numeric

metadata <- tibble("Group" = c("L1", "L2", "S3", "S4", "L5", "L6", "S7", "S8"),
       "Population" = rep(c("Bovra", "Smadalen", "Helin", "Spiterstulen"), each = 2),
       "Size" = rep(c("Large", "Small", "Large", "Small"), each = 2),
       "Elevation" = rep("High", each = 8),
       "Condition" = rep(c("Stable", "Declining", "Declining", "Declining"), each = 2), #difference between "decline" and "decrease"? https://brightspace.ru.nl/d2l/le/content/345973/viewContent/1977818/View
       "Soil" = rep(c("Riverbed", "Riverbed", "Scree", "Rocks"), each = 2),
       "Treatment" = c("Unsalted", "Salted", "Unsalted", "Salted", "Unsalted", "Salted", "Unsalted", "Salted"),
       "Conc_NaCl" = c(0, 50, 0, 50, 0, 50, 0, 50),
       "Conc_unit" = rep("mM", each = 8))

data <- mutate(data_raw,
               "Length_dif" = data_raw$Length_longest_leaf_a - data_raw$Length_longest_leaf_b,
               "Number_dif" = data_raw$Number_leaves_a - data_raw$Number_leaves_b) #calculating differences between before and after treatment

data_complete <- left_join(data, metadata, by = "Group") #create one complete dataframe
```

Now to begin the data analysis, the Shapiro-Wilk values of each group are calculated too see whether the values are normally distributed.

```{r, warning=FALSE, message=FALSE, results='hide'}
names <- c("Wet_weight", "Length_dif", "Number_dif") #collect the names of the columns of interest

sw_results <- vector("list", 3) #prepare an empty list

for (i in 1:3) { #create a list of dataframes where every dataframe contains its Shapiro-Wilk p-values
result <- data_complete %>% group_by(Size, Treatment) %>% summarize(sw = shapiro.test(!!sym(names[i]))$p.value) %>% rename.vars(to = names[i], from ="sw")
result <- unite(result, "Treatment", Size, Treatment)
 sw_results[[i]] <- result
}

merged <- sw_results %>% reduce(full_join, by= "Treatment") #merge all the dataframes from the list

tidy_sw_results <- pivot_longer(data = merged, cols = names,  
                             names_to = "Info",  values_to = "ShapiroWilk_p.value") #make the dataframe tidy
```

```{r}
tidy_sw_results %>% filter(ShapiroWilk_p.value > 0.05) #filter for which groups are normally distributed
tidy_sw_results %>% filter(ShapiroWilk_p.value < 0.05) #filter for which groups are NOT normally distributed
```

All groups of interest are normally distributed, except for the difference in number of leaves

To check whether there is a difference between the wet weights and growth (difference in length of the longest leaf) between the salted and unsalted groups, a t-test is used after checking for equal variance:

```{r}
leveneTest(data_complete$Wet_weight ~ as.factor(Treatment), data = data_complete) #Pr(>F) = 0.4168, which means equal variance
leveneTest(data_complete$Length_dif ~ as.factor(Treatment), data = data_complete) #Pr(>F) = 0.4793, which means equal variance

t.test(formula = data_complete$Wet_weight ~ data_complete$Treatment, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.011, which means there is a statistical significant difference between the salted and unsalted plants.

data_complete %>%
  group_by(Treatment) %>%
  summarize(mean_leaves.increase = mean(Number_dif, na.rm=TRUE), stdev=sd(Number_dif, na.rm=TRUE))
```

Zooming in, the large population-salted and small population-salted groups will be compared using a t-test for the wet weight and length-increase and a Wilcoxon test for the increase in number of leaves (as this group's values were not normally distributed):

```{r, warning=FALSE}
data_salted <- subset(data_complete, Treatment == "Salted")
leveneTest(data_salted$Wet_weight ~ as.factor(Size), data = data_salted) #Pr(>F) = 0.3754 which means equal variance

t.test(formula = data_salted$Wet_weight ~ data_salted$Size, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.848, which means there is no statistical significant difference between the large and small salted plants.

data_salted <- subset(data_complete, Treatment == "Salted")
leveneTest(data_salted$Length_dif ~ as.factor(Size), data = data_salted) #Pr(>F) = 0.7135 which means equal variance

t.test(formula = data_salted$Length_dif ~ data_salted$Size, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.563, which means there is no statistical significant difference between the large and small salted plants.

wilcox.test(Number_dif ~ Size, data = data_salted) #Wilcoxon test, p = 0.4735, which means there is no statistical significant difference between the difference in number of leaves of the large and small salted plants.
```


