---
title: "Data analysis of Arabidopsis experiment B1"
author: Nine Luijendijk
output: pdf_document
date: "2023-05-26"
---
 <details>
 <summary>Loading the packages</summary>
 
```{r setup, results='hide', message=FALSE}
library(readxl)
library(tidyverse)
library(gdata)
library(car)
library(cowplot)
library(magick)
library(here)
```
</details>

The first step is to load in the data from the Excel file and combine it with the metadata.

<details>
<summary>Loading file and generating complete dataframe</summary>

```{r loading_data}
data_raw <- read_excel("data/arabidopsis_data.xlsx", col_types = c("text", rep("numeric", 8), "text")) #making sure every value is numeric

metadata <- tibble("Group" = c("L1", "L2", "S3", "S4", "L5", "L6", "S7", "S8"),
       "Population" = rep(c("Bovra", "Smadalen", "Helin", "Spiterstulen"), each = 2),
       "Size" = rep(c("Large", "Small", "Large", "Small"), each = 2),
       "Elevation" = rep("High", each = 8),
       "Condition" = rep(c("Stable", "Declining", "Declining", "Declining"), each = 2), #difference between "decline" and "decrease"? https://brightspace.ru.nl/d2l/le/content/345973/viewContent/1977818/View
       "Soil" = rep(c("Riverbed", "Riverbed", "Scree", "Rocks"), each = 2),
       "Treatment" = c("Unsalted", "Salted", "Unsalted", "Salted", "Unsalted", "Salted", "Unsalted", "Salted"),
       "Conc_NaCl" = c(0, 50, 0, 50, 0, 50, 0, 50),
       "Conc_unit" = rep("mM", each = 8),
       "Length_unit" = rep("cm", each = 8),
       "Weight_unit" = rep("grams"), each = 8)

data <- mutate(data_raw,
               "Length_dif" = data_raw$Length_longest_leaf_a - data_raw$Length_longest_leaf_b,
               "Number_dif" = data_raw$Number_leaves_a - data_raw$Number_leaves_b) #calculating differences between before and after treatment

data_complete <- left_join(data, metadata, by = "Group") #create one complete dataframe
```
</details>

Now to begin the data analysis, the Shapiro-Wilk values of each group are calculated too see whether the values are normally distributed.

```{r shapirowilk_1, warning=FALSE, message=FALSE, results='hide'}
names <- c("Wet_weight", "Length_dif", "Number_dif") #collect the names of the columns of interest

sw_results <- vector("list", 3) #prepare an empty list

for (i in 1:3) { #create a list of dataframes where every dataframe contains its Shapiro-Wilk p-values
result <- data_complete %>% group_by(Size, Treatment) %>% summarize(sw = shapiro.test(!!sym(names[i]))$p.value) %>% rename.vars(to = names[i], from ="sw")
result <- unite(result, "Treatment", Size, Treatment)
 sw_results[[i]] <- result
}

merged <- sw_results %>% reduce(full_join, by= "Treatment") #merge all the dataframes from the list

tidy_sw_results <- pivot_longer(data = merged, cols = names,  
                             names_to = "Info",  values_to = "ShapiroWilk_p.value") #make the dataframe tidy

tidy_sw_results %>% filter(ShapiroWilk_p.value > 0.05) #filter for which groups are normally distributed
tidy_sw_results %>% filter(ShapiroWilk_p.value < 0.05) #filter for which groups are NOT normally distributed
```

All groups of interest are normally distributed, except for the difference in number of leaves

To check whether there is a difference between the wet weights and growth (difference in length of the longest leaf) between the salted and unsalted groups, a t-test is used after checking for equal variance:

```{r levene_test1}
leveneTest(data_complete$Wet_weight ~ as.factor(Treatment), data = data_complete) #Pr(>F) = 0.4168, which means equal variance
leveneTest(data_complete$Length_dif ~ as.factor(Treatment), data = data_complete) #Pr(>F) = 0.4793, which means equal variance

t.test(formula = data_complete$Length_dif ~ data_complete$Treatment, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.005, which means there is a statistically significant difference in increase in longest leaf length between the salted and unsalted plants.

t.test(formula = data_complete$Wet_weight ~ data_complete$Treatment, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.011, which means there is a statistically significant difference in wet weight between the salted and unsalted plants.

wilcox.test(Number_dif ~ Treatment, data = data_complete) #Wilcoxon test, p = 0.5845, which means there is no statistically significant difference between the difference in number of leaves of the salted and unsalted plants.
```

These differences are interesting to visualize as they are statistically significant:

<details>
<summary>Generating the first plots</summary>

```{r plotting0}
summary <- data_complete %>%
  group_by(Treatment) %>%
  summarize(mean_noleaves.increase = mean(Number_dif, na.rm = TRUE),
            mean_lengthleaf.increase = mean(Length_dif, na.rm = TRUE),
            mean_wetweight = mean(Wet_weight, na.rm = TRUE),
            stdevLength = sd(Length_dif, na.rm = TRUE),
            stdevWetweight = sd(Wet_weight, na.rm = TRUE),
            stdevNumber = sd(Number_dif, na.rm = TRUE))

plot_lengthdiffull <- summary %>% #plot the increase in longest leaf length grouped by population size
  ggplot(aes(x = Treatment, y = mean_lengthleaf.increase))+
  geom_col(aes(fill = Treatment))+
  geom_errorbar(aes(ymin = mean_lengthleaf.increase - stdevLength,
                    ymax = mean_lengthleaf.increase + stdevLength), width=.2)+
  theme_light()+
  labs(title = "Increase in length of the longest\nleaf, grouped by treatment",
       subtitle = "Error bars depict 1 standard deviation",
       x="Treatment",
       y="Mean increase in the length\nof the longest leaf in cm")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=14),
        plot.subtitle = element_text(size=12))

plot_wetweightfull <- summary %>% #plot the wet weight grouped by population size
  ggplot(aes(x = Treatment, y = mean_wetweight))+
  geom_col(aes(fill = Treatment))+
  geom_errorbar(aes(ymin = mean_wetweight - stdevWetweight,
                    ymax = mean_wetweight + stdevWetweight), width=.2)+
  theme_light()+
  labs(title = "Wet weight, grouped by treatment",
       subtitle = "Error bars depict 1 standard deviation",
       x="Treatment",
       y="Wet weight in grams")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=14),
        plot.subtitle = element_text(size=12))

plot_numberdiffull <- summary %>% #plot the increase in number of leaves grouped by population size
  ggplot(aes(x = Treatment, y = mean_noleaves.increase))+
  geom_col(aes(fill = Treatment))+
  geom_errorbar(aes(ymin = mean_noleaves.increase - stdevNumber,
                    ymax = mean_noleaves.increase + stdevNumber), width=.2)+
  theme_light()+
  labs(title = "Increase in number of leaves,\ngrouped by treatment",
       subtitle = "Error bars depict 1 standard deviation",
       x="Treatment",
       y="Mean increase in the number of leaves")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=14),
        plot.subtitle = element_text(size=12))

plotgrid0 <- plot_grid(plot_lengthdiffull, plot_wetweightfull, plot_numberdiffull, #combine the 3 plots into 1 figure
                       labels = c("A", "B", "C"),
                       ncol = 3, nrow = 1)
```

</details>

```{r, echo=FALSE, fig.width = 10.5, fig.cap = "Figure 1: Bar graphs showing A) the increase in longest leaf length grouped by treatment, B) the increase in number of leaves grouped by treatment, and C) the increase in number of leaves grouped by treatment, with error bars depicting 1 standard deviation."}
plotgrid0
```

Zooming in, the large population-salted and small population-salted groups will be compared using a t-test for the wet weight and length-increase and a Wilcoxon test for the increase in number of leaves (as this group's values were not normally distributed):

```{r wilcox_test1, warning=FALSE}
data_salted <- subset(data_complete, Treatment == "Salted")
leveneTest(data_salted$Wet_weight ~ as.factor(Size), data = data_salted) #Pr(>F) = 0.3754 which means equal variance

t.test(formula = data_salted$Wet_weight ~ data_salted$Size, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.848, which means there is no statistically significant difference between the large and small salted plants.

leveneTest(data_salted$Length_dif ~ as.factor(Size), data = data_salted) #Pr(>F) = 0.7135 which means equal variance

t.test(formula = data_salted$Length_dif ~ data_salted$Size, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.563, which means there is no statistically significant difference between the large and small salted plants.

wilcox.test(Number_dif ~ Size, data = data_salted) #Wilcoxon test, p = 0.4735, which means there is no statistically significant difference between the difference in number of leaves of the large and small salted plants.
```

The data can be plotted, where it's visualized how little difference there is:

 <details>
 <summary>Generating the figures</summary>
```{r plotting1}
summary_salted <- data_salted %>%
  group_by(Size) %>%
  summarize(mean_noleaves.increase = mean(Number_dif, na.rm = TRUE),
            mean_lengthleaves.increase = mean(Length_dif, na.rm = TRUE),
            mean_wetweight = mean(Wet_weight, na.rm = TRUE),
            stdevlengthleave.increase = sd(Length_dif, na.rm = TRUE),
            stdevwetweight = sd(Wet_weight, na.rm = TRUE),
            stdevNumber = sd(Number_dif, na.rm = TRUE))

plot_length <- summary_salted %>% #plot the increase in longest leaf length grouped by population size
  ggplot(aes(x = Size, y = mean_lengthleaves.increase))+
  geom_col(aes(fill = Size))+
  geom_errorbar(aes(ymin = mean_lengthleaves.increase - stdevlengthleave.increase,
                    ymax = mean_lengthleaves.increase + stdevlengthleave.increase), width=.2)+
  theme_light()+
  labs(title = "Increase in length of the longest\nleaf, grouped by population",
       subtitle = "Error bars depict 1 standard deviation,\nplot only showing salted groups",
       x="Population size",
       y="Mean increase in the length\nof the longest leaf in cm")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=14),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#e457b5", "#57e486"))

plot_noleaves <- summary_salted %>% #plot the increase in number of leaves grouped by population
  ggplot(aes(x = Size, y = mean_noleaves.increase))+
  geom_col(aes(fill = Size))+
  geom_errorbar(aes(ymin = mean_noleaves.increase - stdevNumber,
                    ymax = mean_noleaves.increase + stdevNumber), width=.2)+
  theme_light()+
  labs(title = "Increase in number of leaves,\ngrouped by population size",
       subtitle = "Error bars depict 1 standard deviation,\nplot only showing salted groups",
       x="Population size",
       y="Mean increase in the number of leaves")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=15),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#e457b5", "#57e486"))

plot_wetweight <- summary_salted %>% #plot the increase in number of leaves grouped by population
  ggplot(aes(x = Size, y = mean_wetweight))+
  geom_col(aes(fill = Size))+
  geom_errorbar(aes(ymin = mean_wetweight - stdevwetweight,
                    ymax = mean_wetweight + stdevwetweight), width=.2)+
  theme_light()+
  labs(title = "Wet weight of the plants,\ngrouped by population size",
       subtitle = "Error bars depict 1 standard deviation,\nplot only showing salted groups",
       x="Population size",
       y="Mean wet weight in grams")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=15),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#e457b5", "#57e486"))

plotgrid1 <- plot_grid(plot_length, plot_wetweight, plot_noleaves, #combine the 3 plots into 1 figure
          labels = c("A", "B", "C"),
          ncol = 3, nrow = 1)
```

 </details>

```{r plot1, echo=FALSE, fig.width = 10.5, fig.cap = "Figure 2: Bar graphs showing A) the increase in longest leaf length grouped by population size, B) the increase in number of leaves grouped by population size, and C) the increase in number of leaves grouped by population size, with error bars depicting 1 standard deviation, and the data filtered to only show the salted groups."}
plotgrid1
```

Population size is only used as a proxy for inbreeding, but the genotypes of these plants are still unknown. This is why the populations need to be looked at individually as well:

```{r analysis2}
data_salted %>%
  group_by(Population) %>%
  summarise(ShapiroWilk_p.value = shapiro.test(Length_dif)$p.value) #Shapiro-Wilk test, all p-values are larger than 0.05

data_salted %>%
  group_by(Population) %>%
  summarise(ShapiroWilk_p.value = shapiro.test(Wet_weight)$p.value) #Shapiro-Wilk test, all p-values are larger than 0.05

data_salted %>%
  group_by(Population) %>%
  summarise(ShapiroWilk_p.value = shapiro.test(Number_dif)$p.value) #Shapiro-Wilk test, one of the p-values is lower than 0.05

leveneTest(data_salted$Length_dif ~ as.factor(Population), data = data_salted) #Pr(>F) = 0.8361, which means equal variance

leveneTest(data_salted$Wet_weight ~ as.factor(Population), data = data_salted) #Pr(>F) = 0.7846, which means equal variance

aov(Length_dif ~ Population, data_salted) %>% summary.aov() #Pr(>F) = 0.478, which means there is no statistically significant difference

aov(Wet_weight ~ Population, data_salted) %>% summary.aov() #Pr(>F) = 0.961, which means there is no statistically significant difference

kruskal.test(Number_dif ~ Population, data = data_salted) #Kruskal-Wallis test, p = 0.03029, which means there is a statistically significant difference

summary_complete_pop <- data_salted %>%
  group_by(Population) %>%
  summarize(mean_noleaves.increase = mean(Number_dif, na.rm = TRUE),
            mean_lengthleaves.increase = mean(Length_dif, na.rm = TRUE),
            mean_wetweight = mean(Wet_weight, na.rm = TRUE),
            stdevlengthleave.increase = sd(Length_dif, na.rm = TRUE),
            stdevwetweight = sd(Wet_weight, na.rm = TRUE),
            stdevNumber = sd(Number_dif, na.rm = TRUE))
```

And again, these can be plotted:

 <details>
 <summary>Plotting more figures</summary>

```{r plotting2}
plot_length2 <- summary_complete_pop %>% #plot the increase in longest leaf length grouped by population
  ggplot(aes(x = Population, y = mean_lengthleaves.increase))+
  geom_col(aes(fill = Population), color = c("#e457b5", "#e457b5", "#57e486", "#57e486"), linewidth = 2)+
  geom_errorbar(aes(ymin = mean_lengthleaves.increase - stdevlengthleave.increase,
                    ymax = mean_lengthleaves.increase + stdevlengthleave.increase), width=.2)+
  theme_light()+
  labs(title = "Increase in length of the longest\nleaf, grouped by population",
       subtitle = "Error bars depict 1 standard deviation,\nplot only showing salted groups",
       x="Population",
       y="Mean increase in the length\nof the longest leaf in cm")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=14),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#8111ee", "#7eee11", "#f0860f", "#0f79f0"))

plot_noleaves2 <- summary_complete_pop %>% #plot the increase in number of leaves grouped by population
  ggplot(aes(x = Population, y = mean_noleaves.increase))+
  geom_col(aes(fill = Population), color = c("#e457b5", "#e457b5", "#57e486", "#57e486"), linewidth = 2)+
  geom_errorbar(aes(ymin = mean_noleaves.increase - stdevNumber,
                    ymax = mean_noleaves.increase + stdevNumber), width=.2)+
  theme_light()+
  labs(title = "Increase in number of leaves,\ngrouped by population",
       subtitle = "Error bars depict 1 standard deviation,\nplot only showing salted groups",
       x="Population",
       y="Mean increase in the number of leaves")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=15),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#8111ee", "#7eee11", "#f0860f", "#0f79f0"))

plot_wetweight2 <- summary_complete_pop %>% #plot the increase in number of leaves grouped by population
  ggplot(aes(x = Population, y = mean_wetweight))+
  geom_col(aes(fill = Population), color = c("#e457b5", "#e457b5", "#57e486", "#57e486"), linewidth = 2)+
  geom_errorbar(aes(ymin = mean_wetweight - stdevwetweight,
                    ymax = mean_wetweight + stdevwetweight), width=.2)+
  theme_light()+
  labs(title = "Wet weight of the plants,\ngrouped by population",
       subtitle = "Error bars depict 1 standard deviation,\nplot only showing salted groups",
       x="Population",
       y="Mean wet weight in grams")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=15),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#8111ee", "#7eee11", "#f0860f", "#0f79f0"))

plotgrid2 <- plot_grid(plot_length2, plot_wetweight2, plot_noleaves2, #combine the 3 plots into 1 figure
          labels = c("A", "B", "C"),
          ncol = 3, nrow = 1)
```
 
 </details>

```{r plot2, echo=FALSE, fig.width = 10.5, fig.cap = "Figure 3: Bar graphs showing A) the increase in longest leaf length grouped by population, B) the increase in number of leaves grouped by population, and C) the increase in number of leaves grouped by population, with error bars depicting 1 standard deviation and the bar borders showing whether the population is large (pink) or small (green). The data is filtered to only show the salted groups."}
plotgrid2
```

Different types of phenotypes were scored, like the color and shape of the leaves. These scores can be visualized as well:

 <details>
 <summary>Generating even more figures</summary>

```{r plotting3}
data_color <- data_complete %>% #prepare color data for plotting
  mutate(Color_graph = case_when(
    Color == 1 ~ "Green",
    Color == 2 ~ "Yellow",
    Color == 3 ~ "Red",
    Color == 4 ~ "Red and\nyellow"))

plot_color <- data_color %>% #plot number of leaves of each color, grouped by treatment
  ggplot()+ 
  geom_bar(aes(x = Color_graph, fill = Treatment), position = position_dodge2(width = 0.9, preserve = "single"))+
  theme_light()+
  labs(title = "Number of leaves of each color,\ngrouped by treatment",
       x="Color",
       y="Count")

plot_color2 <- data_color %>% #plot number of leaves of each color, grouped by population
  ggplot()+ 
  geom_bar(aes(x = Color_graph, fill = Population, color = Size), linewidth = 1,
           position = position_dodge2(width = 0.9, preserve = "single"))+
  theme_light()+
  labs(title = "Number of leaves of each color,\ngrouped by population",
       subtitle = "Plot only showing salted groups",
       x="Color",
       y="Count")+
  scale_fill_manual(values=c("#8111ee", "#7eee11", "#f0860f", "#0f79f0"))+
  scale_color_manual(values = c("#e457b5", "#57e486"))+
  guides(color=guide_legend(title="Population size"))

plot_shape_salt <- data_complete %>% #plot number of leaves of each shape, grouped by treatment
  ggplot()+ 
  geom_bar(aes(x = Leaf_shape, fill = Treatment), position = position_dodge2(width = 0.9, preserve = "single"))+
  theme_light() +
  labs(title = "Number of leaves of each shape,\ngrouped by treatment",
       x="Leaf shape",
       y="Count")

plot_shape_popsize <- data_salted %>% #plot number of leaves of each shape, grouped by population size
  ggplot()+ 
  geom_bar(aes(x = Leaf_shape, fill = Size), position = position_dodge2(width = 0.9, preserve = "single"))+
  theme_light() +
  labs(title = "Number of leaves of each shape,\ngrouped by population size",
       subtitle = "Plot only showing salted groups",
       x="Leaf shape",
       y="Count")+
  scale_fill_manual(values=c("#e457b5", "#57e486"))+
  guides(fill=guide_legend(title="Population size"))

plot_shape_pop <- data_salted %>% #plot number of leaves of each shape, grouped by population
  ggplot()+ 
  geom_bar(aes(x = Leaf_shape, fill = Population, color = Size), linewidth = 2,
           position = position_dodge2(width = 0.9, preserve = "single"))+
  theme_light() +
  labs(title = "Number of leaves of each shape,\ngrouped by population",
       subtitle = "Plot only showing salted groups",
       x="Leaf shape",
       y="Count")+
  scale_fill_manual(values=c("#8111ee", "#7eee11", "#f0860f", "#0f79f0"))+
  scale_color_manual(values = c("#e457b5", "#57e486"))+
  guides(color=guide_legend(title="Population size"))

img_leafshapes <- image_read(here("images/leaf_shapes.png")) %>% image_ggplot() #import the picture showing the different leaf shapes

plotgrid3 <- plot_grid(plot_color, plot_color2, plot_shape_salt, img_leafshapes, plot_shape_popsize, plot_shape_pop, #combine the 3 plots and the leaf shape photo into 1 figure
          labels = c("A", "B", "C", "D", "E", "F"),
          ncol = 2, nrow = 3)
```
 
 </details>
 
```{r plot3, echo=FALSE, fig.height = 13, fig.width = 10.5, fig.cap = "Figure 4: Bar graphs and an image showing A) the number of leaves of each color, grouped by treatment, B) the number of leaves of each color, grouped by population, C) the number of leaves of each shape, grouped by treatment D) the picture showing which shape is equal to which number, E) the number of leaves of each shape, grouped by population size, and F) the number of leaves of each shape, grouped by population. The data is filtered to only show the salted groups in B, E, and F."}
plotgrid3
```

