---
title: "Data analysis of Arabidopsis experiment B1"
author: Nine Luijendijk
output: html_document
date: "2023-05-25"
---

```{r setup, results='hide', message=FALSE}
library(readxl)
library(tidyverse)
library(gdata)
library(car)
library(cowplot)
library(magick)
library(here)
```

The first step is to load in the data from the Excel file and combine it with the metadata.

```{r}
data_raw <- read_excel("data/arabidopsis_data.xlsx", col_types = c("text", rep("numeric", 8), "text")) #making sure every value is numeric

metadata <- tibble("Group" = c("L1", "L2", "S3", "S4", "L5", "L6", "S7", "S8"),
       "Population" = rep(c("Bovra", "Smadalen", "Helin", "Spiterstulen"), each = 2),
       "Size" = rep(c("Large", "Small", "Large", "Small"), each = 2),
       "Elevation" = rep("High", each = 8),
       "Condition" = rep(c("Stable", "Declining", "Declining", "Declining"), each = 2), #difference between "decline" and "decrease"? https://brightspace.ru.nl/d2l/le/content/345973/viewContent/1977818/View
       "Soil" = rep(c("Riverbed", "Riverbed", "Scree", "Rocks"), each = 2),
       "Treatment" = c("Unsalted", "Salted", "Unsalted", "Salted", "Unsalted", "Salted", "Unsalted", "Salted"),
       "Conc_NaCl" = c(0, 50, 0, 50, 0, 50, 0, 50),
       "Conc_unit" = rep("mM", each = 8))

data <- mutate(data_raw,
               "Length_dif" = data_raw$Length_longest_leaf_a - data_raw$Length_longest_leaf_b,
               "Number_dif" = data_raw$Number_leaves_a - data_raw$Number_leaves_b) #calculating differences between before and after treatment

data_complete <- left_join(data, metadata, by = "Group") #create one complete dataframe
```

Now to begin the data analysis, the Shapiro-Wilk values of each group are calculated too see whether the values are normally distributed.

```{r, warning=FALSE, message=FALSE, results='hide'}
names <- c("Wet_weight", "Length_dif", "Number_dif") #collect the names of the columns of interest

sw_results <- vector("list", 3) #prepare an empty list

for (i in 1:3) { #create a list of dataframes where every dataframe contains its Shapiro-Wilk p-values
result <- data_complete %>% group_by(Size, Treatment) %>% summarize(sw = shapiro.test(!!sym(names[i]))$p.value) %>% rename.vars(to = names[i], from ="sw")
result <- unite(result, "Treatment", Size, Treatment)
 sw_results[[i]] <- result
}

merged <- sw_results %>% reduce(full_join, by= "Treatment") #merge all the dataframes from the list

tidy_sw_results <- pivot_longer(data = merged, cols = names,  
                             names_to = "Info",  values_to = "ShapiroWilk_p.value") #make the dataframe tidy
```

```{r}
tidy_sw_results %>% filter(ShapiroWilk_p.value > 0.05) #filter for which groups are normally distributed
tidy_sw_results %>% filter(ShapiroWilk_p.value < 0.05) #filter for which groups are NOT normally distributed
```

All groups of interest are normally distributed, except for the difference in number of leaves

To check whether there is a difference between the wet weights and growth (difference in length of the longest leaf) between the salted and unsalted groups, a t-test is used after checking for equal variance:

```{r}
leveneTest(data_complete$Wet_weight ~ as.factor(Treatment), data = data_complete) #Pr(>F) = 0.4168, which means equal variance
leveneTest(data_complete$Length_dif ~ as.factor(Treatment), data = data_complete) #Pr(>F) = 0.4793, which means equal variance

t.test(formula = data_complete$Wet_weight ~ data_complete$Treatment, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.011, which means there is a statistical significant difference between the salted and unsalted plants.

data_complete %>% group_by(Treatment) %>%
  summarize(mean_noleaves.increase = mean(Number_dif, na.rm = TRUE),
            mean_lengthleave.increase = mean(Length_dif, na.rm = TRUE),
            mean_wetweight = mean(Wet_weight, na.rm = TRUE),
            stdev=sd(Number_dif, na.rm = TRUE))
```

Zooming in, the large population-salted and small population-salted groups will be compared using a t-test for the wet weight and length-increase and a Wilcoxon test for the increase in number of leaves (as this group's values were not normally distributed):

```{r, warning=FALSE}
data_salted <- subset(data_complete, Treatment == "Salted")
leveneTest(data_salted$Wet_weight ~ as.factor(Size), data = data_salted) #Pr(>F) = 0.3754 which means equal variance

t.test(formula = data_salted$Wet_weight ~ data_salted$Size, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.848, which means there is no statistical significant difference between the large and small salted plants.

data_salted <- subset(data_complete, Treatment == "Salted")
leveneTest(data_salted$Length_dif ~ as.factor(Size), data = data_salted) #Pr(>F) = 0.7135 which means equal variance

t.test(formula = data_salted$Length_dif ~ data_salted$Size, 
       paired = FALSE, var.equal = TRUE)$p.value %>% round(.,3) #t-test, p = 0.563, which means there is no statistical significant difference between the large and small salted plants.

wilcox.test(Number_dif ~ Size, data = data_salted) #Wilcoxon test, p = 0.4735, which means there is no statistical significant difference between the difference in number of leaves of the large and small salted plants.
```

The data can be plotted, where it's visualized how little difference there is:

```{r, fig.width = 10, fig.cap = "Figure 1: Bar graphs showing A) the increase in longest leaf length grouped by population size, B) the increase in number of leaves grouped by population size, and C) the increase in number of leaves grouped by treatment, with error bars depicting 1 standard deviation."}
summary_salted <- data_salted %>%
  group_by(Size) %>%
  summarize(mean_noleaves.increase = mean(Number_dif, na.rm = TRUE),
            mean_lengthleaves.increase = mean(Length_dif, na.rm = TRUE),
            mean_wetweight = mean(Wet_weight, na.rm = TRUE),
            stdevlengthleave.increase = sd(Length_dif, na.rm = TRUE),
            stdevwetweight = sd(Wet_weight, na.rm = TRUE),
            stdevNumber = sd(Number_dif, na.rm = TRUE))

plot_length <- summary_salted %>% #plot the increase in longest leaf length grouped by population size
  ggplot(aes(x = Size, y = mean_lengthleaves.increase))+
  geom_col(aes(fill = Size))+
  geom_errorbar(aes(ymin = mean_lengthleaves.increase - stdevlengthleave.increase,
                    ymax = mean_lengthleaves.increase + stdevlengthleave.increase), width=.2)+
  theme_light()+
  labs(title = "Increase in number of leaves,\ngrouped by population size",
       subtitle = "Error bars depict 1 standard deviation",
       x="Population size",
       y="Mean increase in the length\nof the longest leaf in cm")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=15),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#e457b5", "#57e486"))

plot_noleaves <- summary_salted %>% #plot the increase in number of leaves grouped by population size
  ggplot(aes(x = Size, y = mean_noleaves.increase))+
  geom_col(aes(fill = Size))+
  geom_errorbar(aes(ymin = mean_noleaves.increase - stdevNumber, ymax = mean_noleaves.increase + stdevNumber), width=.2)+
  theme_light()+
  labs(title = "Increase in number of leaves,\ngrouped by population size",
       subtitle = "Error bars depict 1 standard deviation",
       x="Population size",
       y="Mean increase in the number of leaves")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=15),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#e457b5", "#57e486"))

plot_wetweight <- summary_salted %>% #plot the increase in number of leaves grouped by treatment
  ggplot(aes(x = Size, y = mean_wetweight))+
  geom_col(aes(fill = Size))+
  geom_errorbar(aes(ymin = mean_wetweight - stdevwetweight, ymax = mean_wetweight + stdevwetweight), width=.2)+
  theme_light()+
  labs(title = "Wet weight of the plants,\ngrouped by population size",
       subtitle = "Error bars depict 1 standard deviation",
       x="Population size",
       y="Mean wet weight in grams")+
  theme(legend.position = "none", text = element_text(size=12), plot.title = element_text(size=15),
        plot.subtitle = element_text(size=12))+
  scale_fill_manual(values=c("#e457b5", "#57e486"))

plot_grid(plot_length, plot_wetweight, plot_noleaves, #combine the 3 plots into 1 figure
          labels = c("A", "B", "C"),
          ncol = 3, nrow = 1)
```

Different types of phenotypes were scored, like the color and shape of the leaves. These scores can be visualized as well:

```{r, fig.width = 10, fig.cap = "Figure 2: Bar graphs and an image showing A) the number of leaves of each color, grouped by treatment, B) the picture showing which shape is equal to which number, C) the number of leaves of each shape, grouped by treatment and D), the number of leaves of each shape, grouped by population size."}
data_color <- data_complete %>% #prepare color data for plotting
  mutate(Color_graph = case_when(
    Color == 1 ~ "Green",
    Color == 2 ~ "Yellow",
    Color == 3 ~ "Red",
    Color == 4 ~ "Red and\nyellow"))

plot_color <- data_color %>% #plot the number of leaves of each color, grouped by treatment
  ggplot()+ 
  geom_bar(aes(x = Color_graph, fill = Treatment), position = position_dodge2(width = 0.9, preserve = "single"))+
  theme_light()+
  labs(title = "Number of leaves of each color,\ngrouped by treatment",
       x="Color",
       y="Count")

plot_shape_salt <- data_complete %>% #plot the number of leaves of each shape, grouped by treatment
  ggplot()+ 
  geom_bar(aes(x = Leaf_shape, fill = Treatment), position = position_dodge2(width = 0.9, preserve = "single"))+
  theme_light() +
  labs(title = "Number of leaves of each shape,\ngrouped by treatment",
       x="Leaf shape",
       y="Countt")

plot_shape_pop <- data_complete %>% #plot the number of leaves of each shape, grouped by population size
  ggplot()+ 
  geom_bar(aes(x = Leaf_shape, fill = Size), position = position_dodge2(width = 0.9, preserve = "single"))+
  theme_light() +
  labs(title = "Number of leaves of each shape,\ngrouped by population size",
       x="Leaf shape",
       y="Countt")

img_leafshapes <- image_read(here("images/leaf_shapes.png")) %>% image_ggplot() #import the picture showing the different leaf shapes

plot_grid(plot_color, img_leafshapes, plot_shape_salt, plot_shape_pop, #combine the 3 plots and the leaf shape photo into 1
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```

